### 2020-07-09 manuscript/manuscript.rmd
updated revisions from Xingxing and Antonis

^2^Department of Agriculture, Zhejiang University, Hangzhou, Zhejiang, China

@@

Yuanning Li: 0000-0001-5765-1419
Xing-Xing Shen: 0000-0002-8436-595X
Benjamin Evans:
Steven H.D. Haddock: 0000-0001-9420-4482
Antonis Rokas: 0000-0002-7248-6551
Casey W. Dunn: 0000-0003-0628-5150

@@

There has been considerable recent debate about the placement of the root in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic analyses and performing new standardized analyses. Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Our new analyses are consistent with this. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories. Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1D) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1C). Since then, many more relevant phylogenomic studies have been published, with some analyses that find support for Ctenophora-sister, some for Porifera-sister, and some neither (Extended Data Table 1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig. 1E). But the extensive technical variation has also made it difficult to synthesize these results to understand the details of this sensitivity. This methodological variance is often confounded, for example when a particular model is only available in one software package.

@@

The models in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR). The analyses considered here have site homogenous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices. Model effect is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). The composition of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

taxon_rectangles$matrix =
  factor(
    taxon_rectangles$matrix,
    levels = c("Dunn2008", "Hejnol2009", "Philippe2009", "Nosenko2013_ribo_14615", "Ryan2013_est",
             "Moroz2014_3d", "Whelan2015_D20", "Borowiec2015_Total1080", "Chang2015",
             "Whelan2017_full", "Simion2017"))

@@

Ingroup taxon sampling also varies widely between studies (Fig. 2B, Extended Data Fig. 1). Sampling of ingroup taxa (animals) was initially biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 1). This variation is in part because newer sequencing technologies in more recent papers are usually not applied to the exact same species that were in earlier papers.

@@

**Fig. 3.**  A summary of phylogenomic results from the previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

@@

Our summary of previous phylogenetic analyses (Fig. 3A) show that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem (XX), but draws on a more systematic overview. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation can make it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity across sites.

@@

Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb]. For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()`from 11 studies.

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) get different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. CAT models employ a Dirichlet process prior to infer the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from C10 to C60. Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This requires a very large number of additional parameters that must be estimated from the data. Despite C60 models have a better fit than site-homogeneous mdoels in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probabaly already too high according to ModelFinder (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories. When we swept the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.

@@

To further explore the variation described above, we quantify the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 3). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models. Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to week Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4 when outgroups are restricted (Choanimalia, Supplementary Table 7).

The fact that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.

@@

## The current state of understanding and future direction

@@

Overcoming the challenges that face this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context. In particular, there are very few chromosome-level genome assemblies outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

For Poisson+CAT models, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. [XX Place holder for Xingxing]

@@

### S.1 Models of molecular evolution
Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. Papers often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT).  Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used.

@@

Note the main conclusion of this paper is baed on the selection of genes that evolved slowly ribosomal RNA genes. The Ctenophora-sister was rejected based on the most saturated dataset and not supported by the matrix with the most closely related outgroups (Choanomalia matrix) [@nosenko2013deep].

@@

Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprised 1080 orthologs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included. The orthologs were constructed using OrthologID pipeline [@chiu2006orthologid]. After removal of spurious sequences and genes with more than 40% of mission data, the final matrix included 1080  (Total 1080). The authors further filtered the full dataset to 9 sub-datasets by filtering genes with high long-branch scores; genes with high saturation; gene occupancy; fast evolving genes. The main conclusion of the paper was largely based on BorowiecTotal_1080 and Borowiec_Best108 matrices. Phylogenetic analyses were conducted under the GTR+CAT model in PhyloBayes in selected matrices, and under the data-partitioning methods in RAxML for all matrices.

@@

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015, with the difference of a core Ctenophora core datasets were constructed here with more than 2000 genes. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset. The main conclusions of this paper were based on Whelan2017_full and Whelan2017_strict matrices.

>>>>

^2^2State Key Laboratory of Rice Biology and Ministry of Agriculture Key Lab of Molecular Biology of Crop Pathogens and Insects, Institute of Insect Sciences, Zhejiang University, Hangzhou 310058, China

@@

Yuanning Li: 0000-0002-2206-5804
Xing-Xing Shen: 0000-0002-8436-595X
Benjamin Evans:
Steven H.D. Haddock: 0000-0001-9420-4482
Antonis Rokas: 0000-0002-7248-6551
Casey W. Dunn: 0000-0003-0628-5150

@@

There has been considerable debate about the placement of the root in the animal tree of life. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). [XXXIt would be helpful if you also stated why this is a problem: Antonis] These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses. Published studies find support for Porifera-sister in some analyses with reduced outgroup (non-animal) sampling and complex site heterogeneous models, and Ctenophora-sister under other conditions. Our new analyses are consistent with this. Though this had been interpreted by some to suggest Ctenophora-sister is an artifact of amino-acid composition bias, our analyses instead indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of substitutional categories. [XXXTo me this part is a diversion – I would state your key inference (“our analyses indicate that many site heterogeneous models do support Ctenophora-sister, and Porifera-sister is a result of unconstrained overfit models with a very high number of site categories”) upfront. You can then explain how your results fit / shed light on previous interpretations of certain analyses.]Our analyses help reconcile variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1E) was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1D). Since then, many more relevant phylogenomic studies have been published, with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, revealing for example that outgroup and model selection can have a large impact on result (Fig. 1). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. This methodological variance is often confounded, for example when a particular model is only available in one software package.

@@

The models in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR). The analyses considered here have site homogenous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices. Model effect is further confused by the abbreviations that are used for models. studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). The composition of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

taxon_rectangles$matrix =
  factor(
    taxon_rectangles$matrix,
    levels = c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_ribo_14615", "Ryan2013_est",
             "Moroz2014_3d", "Borowiec2015_Total1080", "Whelan2015_D20", "Chang2015",
             "Simion2017", "Whelan2017_full"))

@@

Ingroup taxon sampling also varies widely between studies (Fig. 2B, Extended Data Fig. 1). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 1). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were in earlier studies.

@@

**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parametersto the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

@@

Our summary of previous phylogenetic analyses (Fig. 3A) show that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge and thus renders the analysis untrustworthy / hard to interpret. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation makes it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity in equilibrium amino acid frequencies across sites.

@@

Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb]. For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()`data matrices from 11 studies.

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. CAT models employ a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60). Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr %>% mean() %>% round(1)` categories for GTR+CAT analyses. This requires a very large number of additional parameters that must be estimated from the data. Despite C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probably already too high according to ModelFinder (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories. When we varied the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.

@@

To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 3). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE. Moreover, the outgroup choice has little impact on the distribution of the support of phylogenetic signal in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to week Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4 when outgroups are restricted (Choanimalia, Supplementary Table 7).

The observation that support across genes for Ctenophora-sister is not sensitive to outgroup sampling under site homogenous models, but is sensitive under site heterogeneous models, indicates that XX.

@@

## The current state of understanding and future directions

@@

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl). For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300). Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study15

@@

### S.1 Models of molecular evolution
Models of molecular evolution are assembled by selecting different options for all these different components. The models that are applied in practice are heavily influenced by computational costs, as well as convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices. LG and WAG exchangeability matrices have only been used with site homogeneous estimates of equilibrium frequency. This is further confused by the abbreviations that are used for models. studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to equilibrium frequency variation across sites and WAG a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. One could, though, run a WAG+CAT model, although the computation time is much longer than Poisson+CAT model. To avoid confusion on this point, we always specify the exchangeability matrix first, followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.*, CAT).  Moreover, the Gamma-rate heterogeinity (G4) is used in almost every analysis conducted here. If there are no modifiers, then it is implied that site homogeneous models are used.

@@

Note the main conclusion of this study is baed on the selection of genes that evolved slowly ribosomal RNA genes. The Ctenophora-sister was rejected based on the most saturated dataset and not supported by the matrix with the most closely related outgroups (Choanomalia matrix) [@nosenko2013deep].

@@

Borowiec *et al.* [@borowiec2015extracting] assembled a genome dataset comprised 1080 orthologs derived from 36 publicly available genomes representing major lineages of animals, although only one genome of sponge and ctenophore was included. The orthologs were constructed using OrthologID pipeline [@chiu2006orthologid]. After removal of spurious sequences and genes with more than 40% of mission data, the final matrix included 1080  (Total 1080). The authors further filtered the full dataset to 9 sub-datasets by filtering genes with high long-branch scores; genes with high saturation; gene occupancy; fast evolving genes. The main conclusion of the study was largely based on BorowiecTotal_1080 and Borowiec_Best108 matrices. Phylogenetic analyses were conducted under the GTR+CAT model in PhyloBayes in selected matrices, and under the data-partitioning methods in RAxML for all matrices.

@@

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015, with the difference of a core Ctenophora core datasets were constructed here with more than 2000 genes. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset. The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.
